Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-05-10 paper1_jiturra.Rmd
actualizacion intro y seccion1

\pagebreak 
# Introduction

The study of redistributive preferences and social class has been mainly focused on the individual and household situation. However, less attention has been paid to the sociability structures embedded in class relations regarding social tie formation. Moreover, friendship and partnership ties tend to be homogeneous due to the homophilous tendency in interpersonal network formation [@mcpherson_birds_2001;@lazarsfeld_friendship_1954]. Thus, the scope of class analysis in attitude formation urges to be extended to stress the role of social ties and segregation as they involve social processes, such as class identity formation, internalization of shared norms, conflict, and cooperation between classes [@Svallfors2006]. In this respect, recent research has argued that the degree of segregation in the composition of social ties can influence opinions about economic inequality as they function as a window for learning about the lifestyles, economic conditions, and worldviews of other social classes [@mijs_inequality_2018;@otero_power_2023]. However, a knowledge gap in the literature still needs to be understood regarding the network mechanisms in forming individual opinions about governmental actions that aspire to achieve egalitarian economic distribution of resources and opportunities in society [@mccall_americans_2009].

Recent studies have argued that understanding class relations requires focusing on network composition as it represents cross-class embeddedness as a property of the social structure [@blau_macrosociological_1977]. In this regard, it has been argued that the class position of network ties can influence opinions on redistributive policies, but also the degree of segregation from other social classes in more homogeneous social networks can strengthen the class divide in redistributive preferences  [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022]. While prior research has predominantly focused on examining the impact of social class through an individualistic lens, it is noteworthy that more attention must be devoted to understanding the role of social environments in class relations. This omission is particularly surprising given that class positions are fundamentally rooted in production relations that make them inherently relational, not only in their economic underpinnings but also in the power dynamics entwined within class conflicts [@wright_comparative_1989]. In addition, the normative basis of class relations introduces the relevance of the dimensions of solidarity and reciprocity, which have been argued to provide the moral basis for legitimacy and popular support for welfare schemes [@mau_moral_2003].

Comparative evidence about the contextual factors that affect the relationship between class-based network segregation and redistributive preferences is scarce. Previous research on the class-attitude link has suggested that the distribution of economic resources is one of the main societal characteristics that can affect the link between class relations and redistributive demands, as it represents the current state of the distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. These studies have indicated that the affluent are more sensitive to changes in income inequality and support redistribution to a greater extent when inequality rises, while the working class remains relatively stable. In this discussion, the sociological moral economy perspective has suggested that the upper classes are more concerned about the erosion of the opportunity structure as a matter of distributive justice [@sachweh_why_2019]. In contrast, the political economy approach has suggested that the negative consequences of inequality (e.g., criminality) can motivate greater support for redistribution among the upper classes [@dimick_altruistic_2017].

It has been previously established that wider economic gaps lead to segregated forms of social participation and status composition of social networks [@pichler_patterns_2007;@otero_space_2022]. Indeed, it has been suggested that material inequality reinforces the stratified access to social activities and the social distance between social classes, resulting in an increasing marginalization of people with lower resources and consolidating the privileged positions of the upper classes that held better opportunities for diversifying their social activities and interpersonal ties [@lancee_income_2012;@letki_getting_2015;@otero_differences_2023].

Only a few studies have examined the role of network structure from a comparative perspective, employing a well-established instrument as the position generator [@hadler_issp_2020;@sapin_issp_2020]. Despite previous evidence, to our knowledge, no research has stressed the role of economic inequality in the relationship between network segregation and redistributive preferences from a cross-national perspective.

Therefore, our study is guided by two key research questions. First, at the micro-level, we explore the role of class-based network segregation in shaping redistributive preferences. Secondly, at the macro level, we investigate whether economic inequality conditions the relationship between class-based network segregation and redistributive preferences. To address these questions, we have conducted an analysis of data from the International Social Survey Program (ISSP) 2017, which includes a sample of 32,529 individuals across 31 societies. The ISSP dataset is particularly valuable for our study as it provides comprehensive information on interpersonal networks, social class, and attitudes toward redistribution.

@@

The class cleavage has not been strongly supported while the class differences have been evidenced in perceived class conflicts (Edlund & Lindh, 2015; Pérez, 2023), attitudes on market-based inequality (Andersen &Yaish, 2018; Lindh, 2015), and redistributive preferences (Brooks & Svallfors, 2010; Curtis & Andersen, 2015; Langsæther & Evans, 2020).

@@

As social relations within the occupational structure serve as the foundation for class analysis, widening our examination to other dimensions of social life, such as family dynamics, the development of friendships, and diverse social activities, can significantly enrich the scope of class analysis, particularly concerning its relevance in political attitudes formation. In principle, social stratification scholars have traditionally considered differentiation as observable patterns of association given by the intersection of status and group membership, which can be understood as relational networks between positions [@blau_macrosociological_1977]. Empirically, homophily is one of the most significant findings in network research, understood as the higher likelihood of tie formation between people of similar characteristics [@mcpherson_birds_2001, p.416] that trace friendship and family ties, such as age and gender, as well as segregated social environments alongside ethnicity and socioeconomic status [@bargsted_pautas_2020; @plaza_juntos_2022;@lazarsfeld_friendship_1954]. In addition, psychosocial factors such as socialization preferences and group dynamics also play a role in segregation processes, considering that isolated social environments decrease the interchange of worldviews, reinforcing in-group selectivity and value similarity [@visser_attitudes_2004;@homans_human_1951]. Nonetheless, without denying the relevance of socialization preferences, segregation has been predominantly explained through social differentiation in social activities as foci of contact opportunities that ultimately consolidate social networks [@feld_focused_1981; @mcpherson_network_2019]. From this perspective, we argue that class relations represent not only resource-based distinctions but also patterns of sociability and exchange observed in the differentiation of social ties.

How classes are formed also provides insight into social relations and sociability practices. In sociology, classes are understood as the basis for action given their labor market situation, which ultimately shapes their access to opportunities and resources [@weber_economy_1978]. According to the habitus theory, classes incur a series of distinctive practices pivoted in cultural capital as the basis for mutual recognition [@bourdieu_distinction_1984]. From this approach, classes seek to improve their positions through the intergenerational transmission of resources, where similarities in friendship and family choices play a crucial role in reproducing privileged social positions. Additionally, while symbolic resources reinforce status distinctions, access to exclusive social activities in the upper class increases segregation at the expense of excluding lower classes [@bourdieu_reproduccion_1981]. Ultimately, classes consolidate their demographic and cultural identities by mobilizing resources and shared sociability practices, ensuring the intergenerational reproduction of their structural positions [@goldthorpe_sobre_1992]. Hence, the socialization of common lifestyles and shared worldviews facilitates the consolidation of social classes, where access to resources embedded in social networks follows a pattern of accumulation that ultimately leads to social segregation in both the upper and lower bounds of the social structure.

Previous research on social class and tie formation has suggested that class permeability in social networks, understood as the formation of cross-class social ties, can be traced differently across property and authority boundaries. Empirically, evidence from Western industrialized societies has shown that the formation of friendship ties between owners and the working class is significantly less likely, in contrast to the tendency of higher tie formation between supervisors and manual workers. This suggests that the property dimension is much less permeable than authority-based boundaries, given that class interests increase the social distance between proprietors and workers, while the intermediate class position of supervisors and contact frequency with workers make friendship tie formation more likely [@wright_relative_1992]. Similarly, evidence from Chile has shown that the higher permeability of the intermediate classes contrasts with the more homogeneous networks of the working class, suggesting that their limited resources and lower capacity to be socially engaged ultimately result in a lack of social resources that lead to social isolation, whereas the upper class is less permeable and homogenous because of its tendency to self-select as a practice that ultimately seeks to reproduce their privileged status positions [@otero_open_2021].

By contrast, cross-class embeddedness can be described in terms of social activities, generational changes, and the life course. In this regard, @pichler_social_2009 suggest that higher civic engagement in formal organizations increases the chances of bridging with diverse people among the upper class, in contrast to the working class, which tends to be more homogeneous in civic engagement behavior. Similarly, other studies in Europe and South America have shown that the upper and intermediate classes hold increasingly diverse and prestigious social environments than the working classes, where being socially mobile helps to improve both dimensions in tie formation [@cepic_how_2020;@carrascosa_class_2023]. Nevertheless, it does not equalize the weight of the class background compared with those that are intergenerationally stable in the upper class. Additionally, the little longitudinal evidence suggests that networks change following a dynamic of cumulative advantages regarding its composition in terms of prestige and diversity, where upper classes improve in both dimensions, while the lower class shows more stability across the life course [@volker_social_2020].

Altogether, these studies account for class-based network segregation, as the degree of connectedness of an individual to different occupations is meant to represent social resources embedded in social networks vertically in the social structure [@lin_social_2007]. Drawing on this approach, stratification scholars have focused on how social connections are distributed across class structure as a matter of social integration [@blau_macrosociological_1977]. In addition, homogeneity is described as lacking cross-class network ties and is conceptually more proximate to the homophily principle because it is anchored in individual class positions. Simultaneously, _diversity_ is defined as the rate of dissimilar ties or simply the total ties to certain groups that do not necessarily count with a reference position to describe the network composition. Hence, both approaches provide alternatives for studying social class from a network perspective, which has been increasingly discussed in the stratification literature and scholarly discussion about social class and political attitudes.

@@

On the one hand, one approach accentuates the role of beliefs about economic inequality rooted in social comparison processes as an explanation through which people form opinions about redistributive policies [@condon_economic_2020]. This hypothesis can be traced to the studies on class images and perceived class conflicts [@Evans1992;@kelley_class_1995]. Here, the initial argument is that people form their beliefs through the individual, family, friends, and coworkers’ experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of reference groups [@Evans1992, p. 467]. From this perspective, how people infer the social world is linked to the degree of segregation in their immediate social environment, which influences the intensity and character of the information that ultimately shapes their perceptions of inequality [@mijs_is_2021]. Thus, experience sharing in conversations with socioeconomically diverse networks has been proven to contribute to the accuracy of the images of income and wealth inequalities compared with people in more isolated discussion networks [@summers_deliberating_2022]. Despite this, we argue that this research field has been more focused on the cognitive dimension of preference formation through inequality perceptions, either relying on surveys [e.g. @cansunar_who_2021;@garcia-castro_perceived_2022] or experimental manipulations [e.g. @Cruces2011;@becker_significant_2021] rather than empirically addressing the role of class segregation in social networks and its claimed influence on attitudes in the economic domain.

@@

In contrast to studies on participation in formal and informal networks, recent research has emphasized economic inequality in the structure of interpersonal networks. For instance, @letki_getting_2015 argued that economic inequality may constrain access to more extensive networks and the chances of receiving support through them. First, they found that individuals in lower-income groups have smaller networks than those in wealthier groups and are less likely to access resources through them. However, as inequality increases, low-income individuals hold much more extensive family networks but are less inclined towards utilizing their social ties to access resources than middle- and higher-income groups. They suggest that "the poor see maintaining ties with a large extended family as crucial for accessing needed help, which they are unlikely to reach through non-kinship ties" [@letki_getting_2015, p. 244]. By contrast, @otero_differences_2023 found that income inequality is not directly associated with the network structure. However, they found that income inequality can enhance the relationship between educational level, occupational status, and the diversity of social ties. They suggest that income inequality can exacerbate social stratification, creating greater interdependence between cultural, economic, and social capital. Hence, they argue that the unique positions held by the upper classes enable them to navigate different social settings while remaining segregated, whereas the lower classes may experience greater marginalization and isolation because of the choices of others [@otero_open_2021, p.24]. In this regard, if we assume that economic inequality sets the conditions for sociability, a possible consequence may be that the degree of segregation in interpersonal networks and its influence on popular views about economic inequality is also affected by the extent of income differences.

@@

Moving to the bivariate macro relationships, Figure \@ref(fig:fig3) depicts the correlation between network homogeneity, redistributive preferences, and income inequality, our main societal characteristic of interest. First, according to @cohen_statistical_1988, for effect size criteria, we observe a medium positive and significant association between network homogeneity and redistributive preferences in Panel A (_r_ = 0.45, _p_ <.001). In concordance with the previously depicted distribution, we know that higher levels of network homogeneity are driven by the highly segregated networks of the working classes; what makes it logical that in countries where network homogeneity is high, this might also reflect greater social isolation among the working classes and consequently drive higher redistributive preferences. Second, Panel B depicts a positive but relatively weak and nonsignificant association between income inequality and network homogeneity (_r_ = 0.28, _p_<.01). In other words, average homogeneity also tends to increase in societies with greater income gaps. However, the macro association only shows the general pattern between countries but does not allow us to look deeper into class differences according to the levels of inequality.

>>>>

# Introduction

The study of redistributive preferences and social class has been mainly focused on the individual and household situation. Despite this, less attention has been paid to the role of social networks as socialization agents of attitudes toward redistribution. It has been stated that friendship and partnership ties tend to be homogeneous in their characteristics, expressing the tendency to form and consolidate homophilous social networks [@mcpherson_birds_2001;@lazarsfeld_friendship_1954]. Thus, the scope of class analysis in attitude formation urges to be extended to stress the role of social ties and segregation as they involve social processes, such as class identity formation, internalization of shared norms, conflict, and cooperation between classes [@Svallfors2006]. In this respect, recent research has argued that the degree of segregation in the composition of social ties can influence opinions about economic inequality as they function as a window for learning about the lifestyles, economic conditions, and worldviews of other social classes [@mijs_inequality_2018;@otero_power_2023]. However, there is still a knowledge gap to disentangle the network mechanisms involved in political attitude formation about the role of government in enacting redistributive policies to strive against economic inequality [@mccall_americans_2009].

The class composition of network ties can influence opinions on redistributive policies, but also the degree of segregation from other social classes in more homogeneous social networks can strengthen the class divide in redistributive preferences  [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022]. While prior research has predominantly focused on examining the impact of social class through an individualistic lens, it is noteworthy that more attention should be devoted to understanding the role of social environments in class relations. This omission is particularly surprising given that class positions are fundamentally rooted in production relations that make them inherently relational, not only in their economic underpinnings but also in the power dynamics entwined within class conflicts [@wright_comparative_1989]. In addition, the normative basis of class relations introduces the relevance of the dimensions of solidarity and reciprocity, which have been argued to provide the moral basis for legitimacy and popular support for welfare schemes [@mau_moral_2003].

Besides the individual dynamics, little is known about how contextual factors affect the relationship between networks and redistributive preferences. Previous studies have shown that wider economic gaps lead to segregated forms of social participation and status composition of social networks [@pichler_patterns_2007;@otero_space_2022]. Indeed, economic inequality reinforces the stratified access to social activities and the social distance between social classes, resulting in an increasing marginalization of the lower classes and consolidating the privileged positions of the upper classes that held better opportunities for diversifying their social activities and interpersonal ties [@lancee_income_2012;@letki_getting_2015;@otero_differences_2023]. Studies on the class-attitude link posit that economic inequality is crucial for understanding how class relations are reflected in redistributive demands as it represents the current state of the distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. The main finding drawn from these studies is that the upper classes tend to hold more egalitarian attitudes in contexts of high inequality, while the attitudes of the working class remain stable regardless of inequality levels. In this discussion, the moral economy explanation argues that the upper classes are more concerned about the integrity of resource allocation procedures as a matter of distributive justice [@sachweh_why_2019]. Conversely, political economy research suggests that the upper classes are more likely to support redistribution policies due to the negative consequences of inequality [@dimick_altruistic_2017].

In this context, to our knowledge, no research has comparatively stressed the role of economic inequality in the relationship between network segregation and redistributive preferences. Therefore, the present study aims to answer two research questions. First, at the micro-level, what is the role of class-based network segregation on redistributive preferences? Secondly, at the macro level, to what extent does economic inequality condition the relationship between class-based network segregation and redistributive preferences? To address these questions, we have analyzed data from the International Social Survey Program (ISSP) 2017, which includes a sample of 32,529 individuals across 31 societies. In this regard, the ISSP presents a nonprecedence opportunity as it provides comprehensive information on interpersonal networks, social class, and attitudes toward redistribution.

@@

## Class divide in redistributive preferences

Over the past decades, the study of redistributive preferences in industrialized societies has consistently demonstrated the importance of social class as a key explanatory factor  [@lindh_class_2020, p. 421]. Specifically, class positions have been shown to significantly influence individual opinions on the government's role in various aspects of the economic sphere, such as reducing income disparities or enhancing opportunities for those in economic distress [@mccall_americans_2009].

However, class-based explanations of redistributive preferences have mainly been, but not exclusively, focused on the individual or household situation. For instance, it has been claimed that redistributive preferences can be explained through the labor market situation, which comprises access to economic resources and risk exposure [@meltzer_rational_1981;@rehm_risks_2009].
Moreover, as material interests might prevail in scarcity, value-driven motivations are a relevant explanatory factor of redistributive preferences [@Kulin2013;@feldman_humanitarian_2001], becoming salient under greater certainty and weaker under material precariousness [@Maldonadoetal2019]. Another group of studies has suggested that given the substantial time workers spend performing their jobs, social relations in the workspace can also imprint normative views that ultimately shape political opinions [@oesch_redrawing_2006]. It is argued that continuous interaction with diverse people that characterize interpersonal services can reinforce egalitarian values [@kitschelt_occupations_2014]. In contrast, the vertical monitoring in managerial occupations and the relevance of autonomy in the case of self-employed can contribute to self-interested and conservative political views [@langsaether_more_2020;@oesch_electoral_2018].

While our understanding of how social class influences public opinion on market-based resource distribution [@andersen_preferences_2018;@lindh_public_2015] and government income redistribution [@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010;@Svallfors2006] is substantial, there remains a significant gap in our knowledge about the impact of class-based social relations from a network perspective on redistributive preferences.

@@

It is possible to argue that class relations represent not only resource-based distinctions but also patterns of sociability and exchange observed in the differentiation of social ties. Class relations can also be understood as the degree of cross-class relationship as a structural characteristic of a society, representing a network of social ties between different class positions [@blau_macrosociological_1977]. Empirically, homophily in social relations is a consistent finding in the social network literature [@mcpherson_birds_2001, p.416]. Here, friendship and family tend to be homogeneous in terms of their demographic characteristics that simultaneously intersect with socioeconomic status and class [@bargsted_pautas_2020; @plaza_juntos_2022;@lazarsfeld_friendship_1954]. Acknowledging that socialization preferences indeed play a role in the formation of segregated networks [@visser_attitudes_2004;@homans_human_1951], we argue that attitudinal similarity is the result of segregated social relations, which in turn are strongly linked to structural processes of differentiation that ultimately demarcate the opportunities in the formation of diverse social ties [@feld_focused_1981].

In the discussion on the class composition of social networks, it is possible to find two approaches. On the one hand, one approach has emphasized the role of diversity as the degree of connectedness to dissimilar occupations as social positions that vertically represent access to resources embedded in social networks [@lin_social_2007]. However, diversity is defined as the rate of dissimilar ties or the total ties to certain groups that do not necessarily count with a reference position to describe the network composition. On the other hand, a second approach has underscored the role of network segregation, which has been described as lacking cross-class network ties and is conceptually more proximate to the homophily principle because it is anchored in individual class positions [@otero_open_2021].

Empirically, studies on network diversity have shown that higher civic engagement in formal organizations increases the chances of bridging with diverse people among the upper class, in contrast to the more homogeneous participation of the working class [@pichler_social_2009]. Similar patterns have been found in terms of the composition of social ties, where the upper and intermediate classes hold increasingly diverse and prestigious social environments than the working classes [@cepic_how_2020;@carrascosa_class_2023]. Also, this segregation pattern holds during the life course, where the upper classes hold increasingly diverse social contacts in contrast to the stable networks of the working classes [@volker_social_2020]. In contrast, studies that have focused on segregation suggest that the property dimension is much less permeable than authority-based boundaries, given that class interests increase the social distance between proprietors and workers, while the intermediate class position of supervisors and contact frequency with workers make friendship tie formation more likely [@wright_relative_1992] Similar studies have shown that the intermediate classes' higher permeability contrasts with the working class's more homogeneous networks, suggesting that their limited resources and lower capacity to be socially engaged ultimately result in a lack of social resources that lead to social segregation. In contrast, the upper class is less permeable and homogeneous because it tends to self-select as a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

@@

On the one hand, one approach accentuates the role of beliefs about economic inequality rooted in social comparison processes as an explanation through which people form opinions about redistributive policies [@condon_economic_2020]. This hypothesis can be traced to the studies on class images and perceived class conflicts [@Evans1992;@kelley_class_1995]. Here, the initial argument is that people form their beliefs through the individual, family, friends, and coworkers’ experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of reference groups [@Evans1992, p. 467]. From this perspective, how people infer the social world is linked to the degree of segregation in their immediate social environment, which influences the intensity and character of the information that ultimately shapes their perceptions of inequality [@mijs_is_2021]. Thus, experience sharing in conversations with socioeconomically diverse networks has been proven to contribute to the accuracy of the images of income and wealth inequalities compared with people in more isolated discussion networks [@summers_deliberating_2022]. Despite this, we argue that this research field has been more focused on the cognitive dimension of preference formation through inequality perceptions, either relying on surveys [@cansunar_who_2021;@garcia-castro_perceived_2022] or experimental manipulations [@Cruces2011;@becker_significant_2021] rather than empirically addressing the role of class segregation in social networks and its claimed influence on attitudes in the economic domain.

@@

In contrast to studies on participation in formal and informal networks, recent research has emphasized economic inequality in the structure of interpersonal networks. For instance, @letki_getting_2015 argued that economic inequality may constrain access to more extensive networks and the chances of receiving support through them. First, they found that individuals in lower-income groups have smaller networks than those in wealthier groups and are less likely to access resources through them. However, as inequality increases, low-income individuals hold much more extensive family networks but are less inclined towards utilizing their social ties to access resources than middle- and higher-income groups. They suggest that "the poor see maintaining ties with a large extended family as crucial for accessing needed help, which they are unlikely to reach through non-kinship ties" [@letki_getting_2015, p. 244]. By contrast, @otero_differences_2023 found that income inequality is not directly associated with the network structure. However, they found that income inequality can enhance the relationship between educational level, occupational status, and the diversity of social ties. They suggest that income inequality can exacerbate social stratification, creating greater interdependence between cultural, economic, and social capital. Hence, they argue that the unique positions held by the upper classes enable them to navigate different social settings while remaining segregated, whereas the lower classes may experience greater marginalization and segregation because of the choices of others [@otero_open_2021, p.24]. In this regard, if we assume that economic inequality sets the conditions for sociability, a possible consequence may be that the degree of segregation in interpersonal networks and its influence on popular views about economic inequality is also affected by the extent of income differences.

@@

Moving to the bivariate macro relationships, Figure \@ref(fig:fig3) depicts the correlation between network homogeneity, redistributive preferences, and income inequality, our main societal characteristic of interest. First, according to @cohen_statistical_1988, for effect size criteria, we observe a medium positive and significant association between network homogeneity and redistributive preferences in Panel A (_r_ = 0.45, _p_ <.001). In concordance with the previously depicted distribution, we know that higher levels of network homogeneity are driven by the highly segregated networks of the working classes; what makes it logical that in countries where network homogeneity is high, this might also reflect greater social segregation among the working classes and consequently drive higher redistributive preferences. Second, Panel B depicts a positive but relatively weak and nonsignificant association between income inequality and network homogeneity (_r_ = 0.28, _p_<.01). In other words, average homogeneity also tends to increase in societies with greater income gaps. However, the macro association only shows the general pattern between countries but does not allow us to look deeper into class differences according to the levels of inequality.
